---
title: "varycoef: An R Package to Model Spatially Varying Coefficients"
author: "Jakob A. Dambon"
date: "October 2019, Updated: August 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      fig.width=7, fig.height=4)
library(knitr)
prop_train <- 0.2
```


## Introduction

With the R package `varycoef` we enable the user to analyze spatial data and in a simple, yet versatile way. The underlying idea are *spatially varying coefficients* (SVC) that extend the linear model 

$$ y_i = x_i^{(1)} \beta_1 + ... + x_i^{(p)} \beta_p + \varepsilon_i$$
by allowing the coefficients $\beta_j, j = 1, ..., p$ to vary over space. That is, we assume the following model: 

$$ y_i = x_i^{(1)} \beta_1(s) + ... + x_i^{(p)} \beta_p(s) + \varepsilon_i$$

In particular, we use so-called *Gaussian processes* (GP) to define the spatial structure of the coefficients. Therefore, our models are called *GP-based SVC models*.

In this article, we will show what SVC models are and how to define them. Afterwards, we give a short and illustrative example with synthetic data and show how to apply the methods provided in `varycoef`. Finally, using the well known data set `meuse` from the package `sp`.

### Disclaimer

The analyses and results in this article are meant to introduce the package `varycoef` and **not** to write the a rigorous statistical analysis of a data set. As this article should make the usage of `varycoef` as simple as possible, we skip over some of the technical or mathematical details and in some cases abuse notation. For a rigorous definition, please refer to the resources below. Further, the model estimation is performed on rather small data sets to ease computation ($n < 200$). We recommend to apply SVC models, particularly with many coefficients, on larger data sets.

### Further References

Our package evolved over time and we present some highlights:

- In [Dambon et al. (2021a)](https://doi.org/10.1016/j.spasta.2020.100470) we introduce the GP-based SVC model and the methodology on how to estimate them. Further, we provide a comparison on synthetic and real world data with other SVC methodologies.

- In [Dambon et al. (2022a)](https://doi.org/10.1186/s41937-021-00080-2) we present an in-depth analysis of Swiss real estate data using GP-based SVC models.

- In [Dambon et al. (2022b)](https://doi.org/10.1080/13658816.2022.2097684) we introduce a variable selection method. This is not covered by this article.

### Preliminaries

Before we start, we want to give some prerequisites that you should know about in order to follow the analysis below. Beside a classical linear regression model, we require the knowledge of:

- spatial data and geostatistics 
- Gaussian processes and Gaussian random fields
- covariance functions and how the range and variance parameter influence them
- maximum likelihood estimation

### Set up

The `varycoef` package is available via [CRAN](https://cran.r-project.org/web/packages/varycoef/index.html) or [Github](https://github.com/jakobdambon/varycoef). Latter one hosts the most recent version that is released to CRAN on a regular base.

```{r install, eval=FALSE}
# install from CRAN
install.packages("varycoef")

# install from Github (make sure that you installed the package "devtools")
devtools::install_github("jakobdambon/varycoef")
```

### Where to find help?

Within the R package, you can use these resources: 

```{r help and vignettes}
# attach package
library(varycoef)

# general package help file
help("varycoef")

# where you find this vignette
vignette("Introduction", package = "varycoef")
```

You can find this article on the Github, too. We continue to add more material and examples  in the future.

## Synthetic Data Example

Let's dive into the analysis of some synthetic data. To ease the visualization, we will work with one-dimensional spatial data, i.e., any location $s$ is from the real line $\mathbb R$. We want to highlight that are package is not restricted to such analysis and that we can analyze spatial data from higher dimensions, i.e., $s \in \mathbb R^d$, where $d \geq 1$. 

### Model and Data

As mentioned before, an SVC model extends the linear model by allowing the coefficients to vary over space. Therefore, we can write:

$$ y_i = x_i^{(1)} \beta_1(s) + ... + x_i^{(p)} \beta_p(s) + \varepsilon_i,$$
for some location $s$ where $\beta_1(s)$n indicates the dependence of a coefficient on the space. The coefficients are defined by Gaussian processes, but we skip over this part for now and take a look at a data set that was sampled using the model above. In `varycoef`, a data set named `SVCdata` is provided:

```{r synthetic data}
str(SVCdata)
help(SVCdata)
# number of observations, number of coefficients
n <- nrow(SVCdata$X); p <- ncol(SVCdata$X)
```

It consists of the response `y`, the model matrix `X`, the locations `locs`, and the usually unknown true coefficients `beta`, error `eps`, and true parameters `true_pars`. The model matrix is of dimension $`r n` \times `r p`$, i.e., we have $`r n`$ observations and $p = `r p`$ coefficients. The first column of `X` is identical to 1 to model the intercept.

We will use `r n*prop_train` observations of the data to train the model and leave the remaining `r n*(1-prop_train)` out as a test sample, i.e.:

```{r synthetic data train and test}
# create data frame
df <- with(SVCdata, data.frame(y = y, x = X[, 2], locs = locs))
set.seed(123)
idTrain <- sort(sample(n, n*prop_train))
df_train <- df[idTrain, ]
df_test <- df[-idTrain, ]
```

### Exploratory Data Analysis

We plot the part of the data that is usually available to us:

```{r synthetic data EDA}
par(mfrow = 1:2)
plot(y ~ x, data = df_train, xlab = "x", ylab = "y", 
     main = "Scatter Plot of Response and Covariate")
plot(y ~ locs, data = df_train, xlab = "s", ylab = "y", 
     main = "Scatter Plot of Response and Locations")
par(mfrow = c(1, 1))
```

We note that there is a clear linear dependency between the covariate and the response. However, there is not a clear spatial structure. We estimate a linear model and analyze the residuals thereof.

```{r synthetic data linear model}
fit_lm <- lm(y ~ x, data = df_train)
coef(fit_lm)
# residual plots
par(mfrow = 1:2)
plot(x = fitted(fit_lm), y = resid(fit_lm), 
     xlab = "Fitted Values", ylab = "Residuals", 
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2, col = "grey")
plot(x = df_train$locs, y = resid(fit_lm), xlab = "Location", ylab = "Residuals", 
     main = "Residuals vs Locations")
abline(h = 0, lty = 2, col = "grey")
par(mfrow = c(1, 1))
```

Discarding the spatial information, we observe no structure within the residuals (Figure above, LHS). However, if we plot the residuals against there location, we clearly see some spatial structure. Therefore, we will apply the SVC model next. 

### SVC Model

To estimate an SVC model, we can use the `SVC_mle()` function almost like the `lm()` function from above. As the name suggests, we use a *maximum likelihood estimation* (MLE) for estimating the parameters. For now, we only focus on the mean coefficients, which we obtain by the `coef()` method.

```{r synthetic data svc model}
fit_svc <- SVC_mle(y ~ x, data = df_train, locs = df_train$locs)
coef(fit_svc)
```

The only additional argument that we have to provide explicitly when calling `SVC_mle()` are the coordinates, i.e., the observation locations. The output is an object of class `r class(fit_svc)`. We can apply most of the methods that exist for `lm` objects to out output, too. For instance methods of `summary()`, `fitted()`, or `resid()`. We give the `summary()` output but do not go into details for now. Further, use the `fitted()` and `residuals()` methods to analyze the SVC model. Contrary to a `fitted()` method for an `lm` object, the output does not only contain the fitted response, but is a `data.frame` that contains the spatial deviations from the mean named `SVC_1`, `SVC_2`, and so on (see section Model Interpretation below), the response `y.pred`, and the respective locations `loc_1`, `loc_2`, etc. In our case, there is only one column since the locations are from a one-dimensional domain. 

```{r methods}
# summary output
summary(fit_svc)
# fitted output
head(fitted(fit_svc))
# residual plots
par(mfrow = 1:2)
plot(x = fitted(fit_svc)$y.pred, y = resid(fit_svc), 
     xlab = "Fitted Values", ylab = "Residuals", 
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2, col = "grey")
plot(x = df_train$locs, y = resid(fit_svc), xlab = "Location", ylab = "Residuals", 
     main = "Residuals vs Locations")
abline(h = 0, lty = 2, col = "grey")
par(mfrow = c(1, 1))
```

Compared to the output and residuals of the linear models, we immediately see that the residuals of the SVC model are smaller in range and do not have a spatial structure. They are both with respect to fitted values and locations distributed around 0 and homoscedastic.

### Comparison of Models

We end this synthetic data example by comparing the linear with the SVC model. We already saw some advantages of the SVC model when investigating the residuals. Now, we take a look at the quality of the model fit, the model interpretation, and the predictive performance. 

#### Model Fit

We can compare the models by the log likelihood, Akaike's or the Bayesian information criterion (AIC and BIC, respectively). Again, the corresponding methods are available:

```{r synthetic data  model fit, echo = FALSE}
kable(data.frame(
  Model = c("linear", "SVC"), 
  `log Likelihood` = round(as.numeric(c(logLik(fit_lm), logLik(fit_svc))), 2), 
  AIC = round(c(AIC(fit_lm), AIC(fit_svc)), 2), 
  BIC = round(c(BIC(fit_lm), BIC(fit_svc)), 2)
))
```

In all four metrics the SVC model outperforms the linear model, i.e., the log likelihood is larger and the two information criteria are smaller.

#### Visualization of Coefficients

While the linear model estimates constant coefficients $\beta_j$, contrary and as the name suggests, the SVC model's coefficients vary over space $\beta_j(s)$. That is why the `fitted()` method does not only return the fitted response, but also the fitted coefficients and at their given locations:

```{r synthetic data fitted}
head(fitted(fit_svc))
```

Therefore, the SVC mentioned above is the sum of the mean value from the method `coef()` which we name $\mu_j$ and the zero-mean spatial deviations from above which we name $\eta_j(s)$, i.e., $\beta_j(s) = \mu_j + \eta_j(s)$. We visualize the coefficients at their respective locations:

```{r synthetic data SVC plot}
mat_coef <- cbind(
  # constant coefficients from lm
  lin1 = coef(fit_lm)[1], 
  lin2 = coef(fit_lm)[2], 
  # SVCs
  svc1 = coef(fit_svc)[1] + fitted(fit_svc)[, 1], 
  svc2 = coef(fit_svc)[2] + fitted(fit_svc)[, 2]
)
matplot(
  x = df_train$locs, 
  y = mat_coef, pch = c(1, 2, 1, 2), col = c(1, 1, 2, 2), 
  xlab = "Location", ylab = "Beta", main = "Estiamted Coefficients")
legend("topright", legend = c("Intercept", "covariate x", "linear model", "SVC model"), 
       pch = c(1, 2, 19, 19), col = c("grey", "grey", "black", "red"))
```

#### Spatial Prediction

We use the entire data set to compute the in- and out-of-sample rooted mean square error (RMSE) for the response for both the linear and SVC model. Here we rely on the `predict()` methods. Further, we can compare the predicted coefficients with the true coefficients provided in `SVCdata`, something that we usually cannot do. 

```{r synthetic data predictive performance}
df_svc_pred <- predict(fit_svc, newdata = df, newlocs = df$locs)
mat_coef_pred <- cbind(
  svc1_pred = coef(fit_svc)[1] + df_svc_pred[, 1], 
  svc2_pred = coef(fit_svc)[2] + df_svc_pred[, 2]
)
matplot(x = df$locs, y = mat_coef_pred, 
        xlab = "Location", ylab = "Beta", 
        main = "Predicted vs Actual Coefficients",
        col = c(1, 2), lty = c(1, 1), type = "l")
points(x = df$locs, y = SVCdata$beta[, 1], col = 1, pch = ".")
points(x = df$locs, y = SVCdata$beta[, 2], col = 2, pch = ".")
legend("topright", legend = c("Intercept", "Covariate", "Actual"), 
       col = c("black", "red", "black"), 
       pch = c(NA, NA, "."), 
       lty = c(1, 1, NA))
```

#### Model Interpretation

The linear model is constant and the same for each location, i.e.: 

$$y = \beta_1 + \beta_2\cdot x + \varepsilon =  `r round(coef(fit_lm)[1], 2)` + `r  round(coef(fit_lm)[2], 2)` \cdot x + \varepsilon$$
Here, the SVC model can be interpreted in a similar way where on average, it is simply the mean coefficient value with some location specific deviation $\eta_j(s)$, i.e.:

$$y = \beta_1(s) + \beta_2(s)\cdot x + \varepsilon = \bigl(`r round(coef(fit_svc)[1], 2)` + \eta_1(s)\bigr) + \bigl(`r  round(coef(fit_svc)[2], 2)`+ \eta_2(s) \bigr)   \cdot x + \varepsilon$$

Say we are interested in a particular position, like:
```{r sample location}
(s_id <- sample(n, 1))
```

The coordinate is $s_{`r s_id`} = `r df$locs[s_id]`$. We can extract the deviations $\eta_j(s)$ and simply add them to the model. We receive the following location specific model.

$$y = \beta_1(s_{`r s_id`}) + \beta_2(s_{`r s_id`})\cdot x + \varepsilon = \bigl(`r round(coef(fit_svc)[1], 2)`  `r round(df_svc_pred[s_id, 1], 2)`\bigr) + \bigl(`r  round(coef(fit_svc)[2], 2)`+ `r round(df_svc_pred[s_id, 2], 2)`\bigr)   \cdot x + \varepsilon = `r round(coef(fit_svc)[1] + df_svc_pred[s_id, 1], 2)` + `r  round(coef(fit_svc)[2] + df_svc_pred[s_id, 2], 2)`\bigr) \cdot x + \varepsilon$$
The remaining comparison of the two model at the given location is as usual. Between the both models we see that the intercept of the linear model is larger and the coefficient of $x$ is smaller than the 
SVC model's intercept and coefficient of $x$, respectively.

#### Predictive Performance

Finally, we compare the prediction errors of the model. We compute the rooted mean squared errors (RMSE) for both models and both the training and testing data.

```{r synthetic data RMSE}
SE_lm <- (predict(fit_lm, newdata = df) - df$y)^2
# df_svc_pred from above
SE_svc <- (df_svc_pred$y.pred - df$y)^2
kable(data.frame(
  model = c("linear", "SVC"), 
  `in-sample RMSE` = round(sqrt(c(mean(SE_lm[idTrain]), mean(SE_svc[idTrain]))), 3),
  `out-of-sample RMSE` = round(sqrt(c(mean(SE_lm[-idTrain]), mean(SE_svc[-idTrain]))), 3)
))
```

We notice a significant difference in both RMSE between the models. On the training data, i.e., the in-sample RMSE, we observe and improvement by more than 50%. This is quite common since the SVC model has a higher flexibility. Therefore, it is very pleasing to see that even on the testing data, i.e., the out-of-sample RMSE, the SVC model still improves the RMSE by almost 30% compared to the linear model. 

## Meuse Data Set Example

We now turn to a real world data set, the `meuse` data from the package `sp`.

```{r meuse intro}
library(sp)
# attach sp and load data
data("meuse")

# documentation
help("meuse")

# overview
summary(meuse)
dim(meuse)
```

Our goal is to model the log `cadmium` measurements using the following independent variables:

- `dist`, i.e. the normalized distance to the river Meuse. 
- `lime`, which is a 2-level factor indicating the presence of lime. 
- `elev`, i.e. the relative elevation above the local river bed.

This provides us a model with "cheap" covariates and we can regress our variable of interest on them.  

```{r meuse data and location of interest}
df_meuse <- meuse[, c("dist", "lime", "elev")]
df_meuse$l_cad <- log(meuse$cadmium)
df_meuse$lime <- as.numeric(as.character(df_meuse$lime))
locs <- as.matrix(meuse[, c("x", "y")])
```

### Exploratory Data Analysis

First, we plot the log Cadmium measurements at their respective locations. The color ranges from yellow (high Cadmium measurements) to black (low Cadmium measurements).

```{r meuse data spatial plot, echo = FALSE, message=FALSE, warning=FALSE}
# load meuse river outlines
data("meuse.riv")
# create spatial object to create spplot
sp_meuse <- df_meuse
coordinates(sp_meuse) <- ~ locs
# visualize log Cadmium measurements along river
spplot(
  sp_meuse, zcol = "l_cad", main = "Meuse River and Log Cadmium Measurements"
  ) + latticeExtra::layer(panel.lines(meuse.riv))
```

Generally, the values for `l_cad` are highest close to the river Meuse. However, there is some spatial structure comparing the center of all observations to the Northern part. Therefore, we expect the `dist` covariate to be an important regressor. Omitting the spatial structure, we can also look at a `pairs` plot.

```{r meuse data pairs}
pairs(df_meuse)
```

Indeed, we note linear relationships between `l_cad` on the one hand and `elev` as well as `dist` on the other hand. Further, when `lime` is equal to 1, i.e., there is lime present in the soil, the Cadmium measurements are higher.

### Linear Model

As a baseline, we start with a linear model:

```{r meuse linear model}
fit_lm <- lm(l_cad ~ ., data = df_meuse)
coef(fit_lm)
```

The residual analysis shows:

```{r LM residuals}
oldpar <- par(mfrow = c(1, 2))
plot(fit_lm, which = 1:2)
par(oldpar)
```

The spatial distribution of the residuals is the following:

```{r LM spatial residuals, echo=FALSE}
# add residuals to spatial object
sp_meuse$res_lm <- resid(fit_lm)
# visualize linear model residuals along river
spplot(sp_meuse, zcol = "res_lm",
     main = "Meuse River and Residuals of Linear Model"
  ) + latticeExtra::layer(panel.lines(meuse.riv))
```

One can observe that there is a spatial structure in the residuals. This motivates us to use an SVC model. 

### SVC Model

The call to estimate the SVC model is again quite similar to the `lm()` call from above. However, we specify further arguments for the MLE using the `control` argument. First, we are using an optimization over the profiled likelihood (`profileLik = TRUE`) and second, we apply parameter scaling (`parscale = TRUE`) in the numeric optimization. Please check the corresponding help file `help("SVC_mle_control")` for more details.

```{r meuse SVC model, warning=FALSE}
fit_svc <- SVC_mle(l_cad ~ ., data = df_meuse, locs = locs, 
                   control = SVC_mle_control(
                     profileLik = TRUE, 
                     parscale = TRUE
                   ))
coef(fit_svc)
```

The obtained mean coefficient values of the linear and SVC model are quite similar, but this does not come as a surprise.

```{r meuse fixed effects, echo = FALSE}
kable(t(data.frame(
  round(cbind(`linear` = coef(fit_lm), `SVC`= coef(fit_svc)), 3)
)))
```

### Predictions

Additionally to the `meuse` data the `sp` package also contains another data set called `meuse.grid` that contains a 40 by 40 meter spaced grid of the entire study area along the Meuse river. We can use the locations to predict the SVCs. However, the covariates `elev` and `lime` are missing. Therefore, we cannot predict the response. Again, the fitted SVC values are only the deviations from the mean. We observe that the SVC for `elev` is in fact constant.

```{r varycoef predict locations}
# study area
data("meuse.grid")
# prediction
df_svc_pred <- predict(fit_svc, newlocs = as.matrix(meuse.grid[, c("x", "y")]))
colnames(df_svc_pred)[1:4] <- c("Intercept", "dist", "lime", "elev")
head(df_svc_pred)
```

The attentive reader might have noticed that the `elev` column only contains 0. Therefore, there are no deviations and the respective coefficient is constant and our method is capable to estimate constant coefficients within the MLE. There exists a selection method to not only select the varying coefficients, but to also select the mean value, i.e., the fixed effect. Please refer to Dambon et al. (2022b) for further information.

## Conclusion

SVC models are a powerful tool to analyze spatial data. With the R package `varycoef`, we provide an accessible and easy way to apply GP-based SVC models that is very close to the classical `lm` experience. If you have further questions or issues, please visit our [Github repository](https://github.com/jakobdambon/varycoef).


## References

Dambon, J. A., Sigrist, F., Furrer, R. (2021) **Maximum likelihood estimation of spatially varying coefficient models for large data with an application to real estate price prediction**, Spatial Statistics, 41 (100470). [doi:10.1016/j.spasta.2020.100470 ](https://doi.org/10.1016/j.spasta.2020.100470)

Dambon, J.A., Fahrländer, S.S., Karlen, S. et al. (2022a) **Examining the vintage effect in hedonic pricing using spatially varying coefficients models: a case study of single-family houses in the Canton of Zurich**, Swiss Journal Economics Statistics 158(2). [doi:10.1186/s41937-021-00080-2](https://doi.org/10.1186/s41937-021-00080-2)

Dambon, J. A., Sigrist, F., Furrer, R. (2022b) **Joint variable selection of both fixed and random effects for Gaussian process-based spatially varying coefficient models**, International Journal of Geographical Information Science [doi:10.1080/13658816.2022.2097684](https://doi.org/10.1080/13658816.2022.2097684)